Understanding meta-analysis through data simulation with applications to power analysis

### 2023-07-07 documents/paper/paper.Rmd
changes in red (addign oberved effect sizes into captions)
fixing typos

As reported in the previous section, beyond the effect size of each study, we need to assign a weight according to the precision. For this reason, we need to calculate the sampling variability of the effect size or the raw measure that will represent the estimation precision. Each raw or standardized effect size measure (e.g., raw mean difference, Cohen's $d$, and Pearson's correlation) has a different formula to calculate the sampling variability. The idea is to choose the appropriate measure considering the study design (e.g., between vs. within-subjects) and available information and find the proper formula to compute the sampling variability. In addition, there are formulas and approaches to convert from one effect size measure to another [@Borenstein2009-mo; @Lipsey2001-vp; @Lakens2013-eq].^[See also the blog post by James Pustejovsky ([link](https://www.jepusto.com/alternative-formulas-for-the-smd/)) for a general method to compute the sampling variability for standardized mean difference measures.] Usually, the effect size sampling variability depends mainly on the sample size that determines the weight assigned during the meta-analytic estimation.

@@

Before diving into the specific simulations, in this section, we define the common aspects of all simulations in the following sections. The current paper focuses on the *two-level*, *fixed,* and *random-effects* models. All the examples refer to primary studies that assess the efficacy of a treatment by comparing a control and an experimental group. \color{red} There are different simulation approaches to meta-analysis concerning how to generate study-level or participant-level data. In simulation studies where the purpose is not evaluating effect size estimators, it is convenient to simulate unstandardized effect size measures (e.g., UMD). Firstly the estimator is unbiased thus not requiring small sample correction [e.g., @Hedges1989-ip; @Hedges1981-za]. Then the effect size and the sampling variance are independent. Similar to the simulation approach by Viechtbauer [@Viechtbauer2005-yh; @Viechtbauer2007-tq] the *experimental* group ($T$) and *control* groups $C$ are sampled from normal distributions respectively $T_i \sim \mathcal{N}(\Delta, 1)$ and $C_i \sim \mathcal{N}(0, 1)$ where $\Delta$ is the unstandardized mean difference (UMD). The UMD and the sampling variance are calculated using Equations \@ref(eq:effsize) and \@ref(eq:effsizevar) where $y$ is the sample estimator of $\Delta$. \color{black}

@@

The most basic model to simulate is the *equal-effects model* (FE). As reported in the previous sections, the *equal-effects model* assumes the presence of a single true effect ($\theta$) and the observed variability is because each effect size is an imprecise estimation of the true effect. In other words, the only source of variability is the sampling variability that depends on the variance of the primary studies (i.e., the sample size). Equations \@ref(eq:equal-effects-model1) and \@ref(eq:equal-effects-model2) depicts the *equal-effects* model.}

@@

sim <- make_data(k = k, nt = n, nc = n, es = theta)
sim <- sim_studies(sim$es, sim$nc, sim$nt, data = sim) # simulated data
res <- rma(yi = yi, vi = vi, method = "FE", data = sim)
```

@@

From a linear regression perspective, both the FE and the RE models can be seen as *intercept-only* models where only the mean (i.e., the linear regression intercept) is estimated Borenstein et al. (2009). As reported in the meta-analysis introduction, the between-study heterogeneity usually represents the true variability of the effect due to differences among primary studies. A natural extension of the *intercept-only* analysis is a model that includes variables (i.e., *moderators*) that could explain the observed heterogeneity among effect sizes. For example, a group of studies could use a particular memory task where the expected effect is higher than another. In this way, considering the
type of task will explain part of the observed heterogeneity as in standard regression models. Figures \@ref(fig:img-metaregression-bin) and \@ref(fig:img-metaregression-num) depict a random-effects meta-regression model for a categorical and numerical predictor.

@@

\begin{align}
\begin{gathered}
y_{iL} = \beta_0 + \delta_i + \beta_1 \times 0 + \epsilon_i \\
d_{iO} = \beta_0 + \delta_i + \beta_1 \times 1 + \epsilon_i
(\#eq:meta-regression-equations)
\end{gathered}
\end{align}

@@

As explained in the introduction, there are several approaches and tools to estimate the power of *fixed* and *random-effects models* [see @Borenstein2009-mo, Chapter 29; @Harrer2019-rl, Chapter 14]. These methods are easy to implement but made strong assumptions, such as the homogeneity of sample size, and did not consider the uncertainty in estimating $\tau^{2}$. Jackson and colleagues [-@Jackson2017-dv]
partially solve the issue by developing an interesting method that takes into account the uncertainty in estimating $\tau^{2}$ without using simulations. However, only Monte Carlo simulations can consider complex scenarios and extra parameters.

@@

@@

(ref:res-equal-effects) \color{red} Summary of the simulated equal-effects model. The only estimated parameter is the average effect* $\theta_{f}$ *with the standard error,
95% confidence interval, and the Wald z-test. The z-test evaluates the null hypothesis that the real effect equals zero.\color{black}

@@

(ref:res-random-effects) \color{red} Summary of the simulated random-effects model: Compared to the equal-effects model, there are more parameters. The $\beta$ is the average effect ($\theta_{r}$) with the standard error, 95% confidence interval, and the Wald z-test. The $\tau^{2}$ is the estimated heterogeneity and the $I^{2}$ (explained in Section \@ref(re-model)) represents the percentage of total variability due to between-study heterogeneity. \color{black}

@@

(ref:img-metaregression-bin) Graphical representation of a *random-effects meta-regression* model with a categorical predictor (Condition A and Condition B). Each grey distribution represents the sampling distribution of included studies. The dotted line is the average effect (i.e., random-effects model without moderators). The effect size differs between conditions A and B, including the *condition* moderator explaining part of the total heterogeneity (red plus green segments). The green segments depict the explained heterogeneity, and the red segments the residual (unexplained) heterogeneity.

>>>>

As reported in the previous section, beyond the effect size of each study, we need to assign a weight according to the precision. For this reason, we need to calculate the sampling variability of the effect size or the raw measure that will represent the estimation precision. Each raw or standardized effect size measure (e.g., raw mean difference, Cohen's $d$, and Pearson's correlation) has a different formula to calculate the sampling variability. The idea is to choose the appropriate measure considering the study design (e.g., between vs. within-subjects) and available information and find the proper formula to compute the sampling variability. In addition, there are formulas and approaches to convert from one effect size measure to another [@Borenstein2009-mo; @Lipsey2001-vp; @Lakens2013-eq].^[See also the blog post by James Pustejovsky ([www.jepusto.com/alternative-formulas-for-the-smd](https://www.jepusto.com/alternative-formulas-for-the-smd/)) for a general method to compute the sampling variability for standardized mean difference measures.] Usually, the effect size sampling variability depends mainly on the sample size that determines the weight assigned during the meta-analytic estimation.

@@

Before diving into the specific simulations, in this section, we define the common aspects of all simulations in the following sections. The current paper focuses on the *two-level*, *equal,* and *random-effects* models. All the examples refer to primary studies that assess the efficacy of a treatment by comparing a control and an experimental group. \color{red} There are different simulation approaches to meta-analysis concerning how to generate study-level or participant-level data. In simulation studies where the purpose is not evaluating effect size estimators, it is convenient to simulate unstandardized effect size measures (e.g., unstandardized mean difference, UMD). Firstly the estimator is unbiased thus not requiring small sample correction [e.g., @Hedges1989-ip; @Hedges1981-za]. Then the effect size and the sampling variance are independent. Similar to the simulation approach by Viechtbauer [@Viechtbauer2005-yh; @Viechtbauer2007-tq] the *experimental* group ($T$) and *control* groups $C$ are sampled from normal distributions respectively $T_i \sim \mathcal{N}(\Delta, 1)$ and $C_i \sim \mathcal{N}(0, 1)$ where $\Delta$ is the unstandardized mean difference (UMD). The UMD and the sampling variance are calculated using Equations \@ref(eq:effsize) and \@ref(eq:effsizevar) where $y$ is the sample estimator of $\Delta$. \color{black}

@@

The most basic model to simulate is the *equal-effects model* (EE). As reported in the previous sections, the *equal-effects model* assumes the presence of a single true effect ($\theta$) and the observed variability is because each effect size is an imprecise estimation of the true effect. In other words, the only source of variability is the sampling variability that depends on the variance of the primary studies (i.e., the sample size). Equations \@ref(eq:equal-effects-model1) and \@ref(eq:equal-effects-model2) depicts the *equal-effects* model.}

@@

sim <- make_data(k = k, nt = n, nc = n, es = theta)
sim <- sim_studies(sim$es, sim$nc, sim$nt, data = sim) # simulated data
res <- rma(yi = yi, vi = vi, method = "EE", data = sim)
```

@@

From a linear regression perspective, both the EE and the RE models can be seen as *intercept-only* models where only the mean (i.e., the linear regression intercept) is estimated Borenstein et al. (2009). As reported in the meta-analysis introduction, the between-study heterogeneity usually represents the true variability of the effect due to differences among primary studies. A natural extension of the *intercept-only* analysis is a model that includes variables (i.e., *moderators*) that could explain the observed heterogeneity among effect sizes. For example, a group of studies could use a particular memory task where the expected effect is higher than another. In this way, considering the
type of task will explain part of the observed heterogeneity as in standard regression models. Figures \@ref(fig:img-metaregression-bin) and \@ref(fig:img-metaregression-num) depict a random-effects meta-regression model for a categorical and numerical predictor.

@@

\begin{align}
\begin{gathered}
y_{iL} = \beta_0 + \delta_i + \beta_1 \times 0 + \epsilon_i \\
y_{iO} = \beta_0 + \delta_i + \beta_1 \times 1 + \epsilon_i
(\#eq:meta-regression-equations)
\end{gathered}
\end{align}

@@

As explained in the introduction, there are several approaches and tools to estimate the power of *equal* and *random-effects models* [see @Borenstein2009-mo, Chapter 29; @Harrer2019-rl, Chapter 14]. These methods are easy to implement but made strong assumptions, such as the homogeneity of sample size, and did not consider the uncertainty in estimating $\tau^{2}$. Jackson and colleagues [-@Jackson2017-dv]
partially solve the issue by developing an interesting method that takes into account the uncertainty in estimating $\tau^{2}$ without using simulations. However, only Monte Carlo simulations can consider complex scenarios and extra parameters.

@@

The supplementary materials can be also found on the Open Science Framework repository (https://osf.io/54djn/)

@@

(ref:res-equal-effects) \color{red} Summary of the simulated equal-effects model. The only estimated parameter is the average effect $\theta$ ($\beta$) with the standard error, 95% confidence interval, and the Wald z-test. The z-test evaluates the null hypothesis that the real effect equals zero.\color{black}

@@

(ref:res-random-effects) \color{red} Summary of the simulated random-effects model: Compared to the equal-effects model, there are more parameters. The $\beta$ is the average effect ($\mu_{\theta}$) with the standard error, 95% confidence interval, and the Wald z-test. The $\tau^{2}$ is the estimated heterogeneity and the $I^{2}$ (explained in Section \@ref(re-model)) represents the percentage of total variability due to between-study heterogeneity. \color{black}

@@

(ref:img-metaregression-bin) Graphical representation of a *random-effects meta-regression* model with a categorical predictor (Condition A and Condition B). Each grey distribution represents the sampling distribution of included studies. The dotted line is the average effect (i.e., random-effects model without moderators). The effect size differs between conditions A and B, including the *condition* moderator explaining part of the total heterogeneity (red plus green segments). The green segments depict the explained heterogeneity, and the red segments the residual (unexplained) heterogeneity. \color{red} The red squares are simulated observed effect sizes from the sampling distribution. \color{black}
